Jahn-Teller effect with rigid octahedral rotations in perovskites

### 2023-09-20 manuscript/paper.tex
a lot of new references added in bib file. gen reviews ref to perovskite added to tex. some imp reviews ref missed before also included.

\com{
Perovskite?s potential applications are varied and include uses in sensors and catalyst electrodes, certain types of fuel cells, solar cells, lasers, memory devices and spintronics applications.
}

 \com{https://www.frontiersin.org/articles/10.3389/fchem.2020.550625/full:\\
advanced functional materials including photocatalysis, electrocatalysts
 high-temperature superconductors, multiferroics, magnetic and magnetoresistive materials, materials for solar cell energy and photoluminescence
}

\com{https://www.perovskite-info.com/introduction:\\
Perovskite materials exhibit intriguing and unusual physical properties that have been extensively studied for both practical applications and theoretical modeling and the materials science and applications of perovskites have been a broad research area open to many revolutionary discoveries for new device concepts. Perovskite?s potential applications are varied and include uses in sensors and catalyst electrodes, certain types of fuel cells, solar cells, lasers, memory devices and spintronics applications.
}

Perovskites  
exhibit a rich variety of intriguing 
electronic, magnetic and optical properties
due to the interplay between
spin, orbital, and structural degrees of freedom
and widely varying electronic correlations.
These materials 
are important
both for practical applications in all sorts of modern devices 
and 
fundamental understanding of various physical phenomena.

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
has also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}
as well.

@@

While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap}
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor~\cite{Goldschmidt1926,KieslichCS15}. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.

>>>>

\com{
%---------------------------------------------------------
% REVIEWS:
WangS03, %REVIEW
%{Epitaxial BiFeO<sub>3</sub> Multiferroic Thin Film Heterostructures}
DagottoS05,
%{Complexity in Strongly Correlated Electronic Systems},
SchlomARMR07,
%Strain Tuning of Ferroelectric Thin Films}
KhomskiiPhys09
%Classifying multiferroics: Mechanisms and effects}
ZubkoARCMP11,
%{Interface Physics in Complex Oxide Heterostructures},
HwangNM12, 
%{Emergent phenomena at oxide interfaces},

BhattacharyaARMR14,
%{Magnetic Oxide Heterostructures},
HellmanRMP17,
%{Interface-induced phenomena in magnetism},
  ChenJPCM17
 % {Charge transfer driven emergent phenomena in oxide heterostructures},
DagottoMRSB08,
  %{Strongly Correlated Electronic Materials: Present and Future},

SalamonRMP01, 
%The physics of manganites: Structure and transport},
DawberRMP05,
% Physics of thin-film ferroelectric oxides},
KosterRMP12,
%Structure, physical properties, and applications of ${\mathrm{SrRuO}}_{3}$ thin films},

%electronic correlations:
~\cite{DagottoS05,DagottoMRSB08}

%applications:
~\cite{SalamonRMP01,
WangS03
DawberRMP05,
SchlomARMR07,
KosterRMP12}

% understanding:
~\cite{ZubkoARCMP11,
HwangNM12, 
BhattacharyaARMR14,
HellmanRMP17,
ChenJPCM17}

%---------------------------------------------------------

LiN19,
% Superconductivity in an infinite-layer nickelate},
LeeN10,
%{A strong ferroelectric ferromagnet created by means of spin--lattice coupling},
BousquetN08,
%{Improper ferroelectricity in perovskite oxide artificial superlattices},

OjaPRL12,
%{${d}^{0}$ Ferromagnetic Interface between Nonmagnetic Perovskites},

WangPRX16,
% {Stabilization of Highly Polar ${\mathrm{BiFeO}}_{3}$-like Structure: A New Interface Design Route for Enhanced Ferroelectricity in Artificial Perovskite Superlattices}
CavigliaPRL12,
%{Ultrafast Strain Engineering in Complex Oxide Heterostructures},
XiaPRB09,
%{Critical thickness for itinerant ferromagnetism in ultrathin films of ${\text{SrRuO}}_{3}$},

BenedekPRL11,
%{Hybrid Improper Ferroelectricity: A Mechanism for Controllable Polarization-Magnetization Coupling},
GarrityPRL14,
% {Hyperferroelectrics: Proper Ferroelectrics with Persistent Polarization},
KobayashiPRL12,
%{Electronic Ferroelectricity in a Molecular Crystal with Large Polarization Directing Antiparallel to Ionic Displacement},

HarrisPRB07,
% {Landau analysis of the symmetry of the magnetic structure and magnetoelectric interaction in multiferroics},
LeePRL10,
%{Epitaxial-Strain-Induced Multiferroicity in ${\mathrm{SrMnO}}_{3}$ from First Principles},
WangPRL15,
%{Observation of Magnetoelectric Multiferroicity in a Cubic Perovskite System: ${\mathrm{LaMn}}_{3}{\mathrm{Cr}}_{4}{\mathrm{O}}_{12}$},

ChenPRL22,
%{Optically Controlled Femtosecond Polariton Switch at Room Temperature}
ChoiPRL10,
% {Cross-Control of Magnetization and Polarization by Electric and Magnetic Fields with Competing Multiferroic and Weak-Ferromagnetic Phases},
KidaPRB08,
%{Electrically driven spin excitation in the ferroelectric magnet ${\text{DyMnO}}_{3}$},
ChoPRL16,
%{Laser Particle Stimulated Emission Microscopy},
deJongPRL12,
%{Coherent Control of the Route of an Ultrafast Magnetic Phase Transition via Low-Amplitude Spin Precession},

ShirakoPRB11,
%{Integer spin-chain antiferromagnetism of the 4$d$ oxide CaRuO${}_{3}$ with post-perovskite structure},
DikinPRL11,
% {Coexistence of Superconductivity and Ferromagnetism in Two Dimensions},
ChoudhuryPRL12,
%{Near-Room-Temperature Colossal Magnetodielectricity and Multiglass Properties in Partially Disordered ${\mathrm{La}}_{2}{\mathrm{NiMnO}}_{6}$},

LeePRL11,
%  {Electrolyte Gate-Controlled Kondo Effect in ${\mathrm{SrTiO}}_{3}$},

Perovskites  
exhibit a rich variety of intriguing 
electronic,
magnetic
 and optical
properties~\cite{ZubkoARCMP11,
HwangNM12, 
BhattacharyaARMR14,
HellmanRMP17,
ChenJPCM17,DagottoS05,DagottoMRSB08}
due to the interplay between
spin, orbital, and structural degrees of 
freedom
and widely varying electronic 
correlations.
These materials 
are important
both for practical applications in all sorts of modern devices 
and 
fundamental understanding of various physical 
phenomena~\cite{SalamonRMP01,
WangS03,
DawberRMP05,
SchlomARMR07,
KosterRMP12}.

%Their crystal structures 
The perovskite structures
have a common formula unit ABX$_3$,
where 
%positive ions A and negative ions X form a cube and an octahedron around positive ion B.
A and B are positive ions while X are negative ions 
%whose six-fold coordination forms octahedra around B.
that form octahedra around  B.
In a cubic perovskite,
the octahedra are symmetric, 
but they can 
distort~\cite{
BersukerCCR75,
KugelSPU82,
MillisN98,
TerakuraPMS07,
HalcrowRSC13,
LufasoAC04}
or simply 
rotate~\cite{
GlazerAC72, ThomasAC96,
WoodwardAC97a,WoodwardAC97b,
HowardAC98,StokesAC02,AngelPRL05}
 to lower the symmetry
and produce, e.g., a rhombohedral, tetragonal or orthorhombic structure.
% cubic perovskite [spacegroup 221] can transform to:
% crystal system: spacegroups
% triclinic: 2
% monoclinic: 11,12,15
% orthorhombic: 59,62,63,71,74
% tetragonal 127,139,140
% rhombohedral/trigonal: 167
% cubic: 204, 221 
Octahedral rotations also occur in other perovskite related structures
and
have been extensively 
studied
due to their crucial role in determining various physical properties,
e.g., in 
ferroelectric~\cite{GhosezNM11,
RondinelliAM12,
SchickPRB14,
SimPRB14,
LeePRL17,
WangPRL22,
ZhouPRB22,
GaoPRB23} and 
multiferroic~\cite{KhomskiiPhys09,
BenedekPRL11,
YangPRL14,
YePRB18,
ZhangPRL20,
ZhangYajunPRL20, 
ZhouPRB21,
ZhangPRL22}
systems.
The interplay of octahedral rotations
and 
spin-orbit coupling~\cite{AmatNL14,KrachJPCM23},
breathing distortions~\cite{BalachandranPRB13},
strain~\cite{Aguado-PuentePRL11},
defects~\cite{JiaPRB22}, and
pressure~\cite{XiangPRB17},
and 
their role in
magnetostructural phase transitions~\cite{JohnsonPRL20},
orbital order~\cite{OgawaPRL12,ChenPRL19}, orbital anisotropy~\cite{HuangPRB21}, 
and
%surface electronic structure under in-plane uniaxial compression~\cite{MoralesPRL23}
surface electronic structure~\cite{MoralesPRL23},
has also been investigated.
% JT interacting with octa tilting:
Since, usually, the octahedra are not just %(Jahn-Teller) 
distorted or rotated but a combination of both,
the influence of these two 
structural mutations %instabilities 
on each other has recently been explored by many authors~\cite{LufasoAC04,CarpenterAC09,ChenPRL19}
as well.

@@

While the tetragonal distortion of the octahedra is associated with
a (cooperative) Jahn-Teller 
effect~\cite{JahnPRSL37,JahnPRSL38,
OpikRS57,
Sturge68,
GehringRPP75,
BersukerCCR75,
KugelSPU82,
MillisN98,
LufasoAC04,
TerakuraPMS07,
HalcrowRSC13,
PavariniChap,
KhomskiiCR21}
(i.e., the orbital degeneracy of the metal ion coordinated by the anionic octahedron
is lifted, which stabilises the distortion),
%[as it distorts the octahedral field at the central site that split its electronic states],
the rotation and tilting of the octahedra are almost always 
explained empirically in terms of 
the relative sizes of the three ions
with the
Goldschmidt tolerance factor~\cite{Goldschmidt1926,KieslichCS15}. 
%[probably because rigid rotation of the octahedra do not change its potential at the central site]
%We only find two exceptions.
